---
title: "COVID-19 Morbidity and Seasonal Trends"
author: "Charles Fraley"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F)
library(tidyverse)
library(forecast)
library(mgcv)
library(reshape)
library(stringi)
library(stringr)
library(knitr)
library(lubridate)
```

# Goal of Analysis

I'll be evaluating if there's a latitude influenced seasonal aspect to COVID mortality rates. This question is influenced by the knowledge that weather and climate has an influence on respiratory illness. If COVID deaths are largely related to respiratory distress, there may be a notable trend.

# Data
<!-- Was the source and a short description of the data provided? -->
## Intake and Source
I'll be using the [time series COVID-19 Data](https://github.com/CSSEGISandData/COVID-19/tree/master/csse_covid_19_data/csse_covid_19_time_series) from the Center for Systems Science and Engineering at Johns Hopkins.
```{r echo=T}
url_in <- "https://github.com/CSSEGISandData/COVID-19/raw/master/csse_covid_19_data/"
time_path <- "csse_covid_19_time_series/time_series_covid19_"
time_filenames <- c(
    "confirmed_US.csv",
    #"confirmed_global.csv",
    #"deaths_global.csv",
    "deaths_US.csv"
)
for (branch_url in time_filenames) {
  time_data <- read.csv(
    url(
      paste(url_in,time_path,branch_url,sep="")
      )
    )
  assign(
    substring(branch_url,1,nchar(branch_url)-4),
    time_data
    )
}
```
# Demonstration of Method

I will be demonstrating my analysis using a single data entry.
1. Get the corresponding rows in confirmed and death. They added
```{r}
working_df <- data.frame(
  date = colnames(deaths_US[-(1:12)]),
  confirmed=as.numeric(
      t(confirmed_US[1,])[-(1:11)]
     ),
  dead=as.numeric(t(deaths_US[1,])[-(1:12)])
)
working_df$date <- mdy(substring(working_df$date,2))
working_df <- mutate(working_df,ratio=dead/confirmed)
```
    - column names: concatenation of the 
    - row names: dates
    - Cells record ratio of $deaths/confirmed$ day by day

```{r eval=T}
full_demo <- ggplot(working_df,aes(x=date,y=ratio))+geom_point()
full_demo
```

Map each area of examination to their latitude
Fit yearly data to $y=C_{periodic\ amplitude}*sin(\frac{2\pi}{1yr} t+C_{offset})+C_{yearly\ average}+C_{trend\ for\ that\ year}*t$

## All Years on Same Axis

1. Strip years from date.
```{r}
working_df <- mutate(working_df,
                     no_year=yday(date)
)
```
```{r}
demoplot <- ggplot(na.omit(working_df),aes(x=no_year,y=ratio))+geom_point()
```
```{r eval=T}
working_df <- na.omit(working_df)
demo_model <- nls(
  ratio ~ a*sin((2*pi*no_year+c)/365)+b, 
  data=working_df,
  start=list(
    a=.001,
    b=.012,
    c=-30
    )
  )
print(summary(demo_model))

working_df$fit <-  predict(demo_model,data=working_df$no_year, interval='confidence')
demoplot <- demoplot+geom_line(aes(no_year,working_df$fit))
# demoplot <- demoplot+geom_ribbon(aes(ymin=lwr,ymax=upr), alpha=0.3)
```
```{r}
demoplot
```

## Each year as separate consideration

1. Separate year as own variable
```{r}
working_df <- mutate(working_df, year=year(date))
```
2. Similar handling except separate by year
```{r}
model2020 <- nls(
  ratio ~ a*sin(2*pi*(no_year+c)/365)+b, 
  data=filter(working_df,year==2020),
  start=list(
    a=.001,
    b=.012,
    c=-30
  )
  )

model2021 <- nls(
  ratio ~ a*sin(2*pi*(no_year+c)/365)+b, 
  data=filter(working_df,year==2021),
  start=list(
    a=.001,
    b=.012,
    c=-30
  )
  )
model2022 <- nls(
  ratio ~ a*sin(2*pi*(no_year+c)/365)+b, 
  data=filter(working_df,year==2022),
  start=list(
    a=.001,
    b=.012,
    c=-30
  )
  )
```

```{r}
working_df$split_fit <-  unlist(
  c(
    predict(
      model2020,
      data=working_df$date, 
      interval='confidence'
    ),
    predict(
      model2021,
      data=working_df$date, 
      interval='confidence'
    ),
    predict(
      model2022,
      data=working_df$date, 
      interval='confidence'
    )
  )
)

split_demo <- ggplot(working_df,aes(x=date,y=ratio))+geom_point()

split_demo <- split_demo+
  geom_line(aes(date,working_df$split_fit))
split_demo
```
# Analysis
For measuring across multiple locations I'll be using the all as one year analysis. Additionally I will be disregarding the ratio values of 0 at the beginning of the pandemic as these are clearly an issue of how data was collected when COVID emerged. From each county, I need the value for periodic amplitude for the model and the latitude.

```{r}
us_agg <- data.frame(
  uid=confirmed_US$UID,
  combined_key=confirmed_US$Combined_Key,
  latitude=confirmed_US$Lat,
  amp=rep(NA,nrow(confirmed_US)),
  amp_err=rep(NA,nrow(confirmed_US)),
  off=rep(NA,nrow(confirmed_US)),
  off_err=rep(NA,nrow(confirmed_US))
)
working_df <- data.frame(
    ydays = yday(mdy(substring(colnames(deaths_US[-(1:12)]),2)))
)
```
```{r}
for (rownum in 1:nrow(us_agg)) {
  if (grepl("Out of",confirmed_US$Admin2[rownum]) || grepl("Unassigned",confirmed_US$Admin2[rownum]) ){
    next
  }
  # print(rownum)
  working_df$ratio <- 
      as.numeric(t(deaths_US[rownum,])[-(1:12)])/
      as.numeric(t(confirmed_US[rownum,])[-(1:11)])
  skip <- F
  tryCatch(
    work_fit <-nls(
    ratio ~ a*sin((2*pi*(ydays+c))/365)+b, 
    working_df[
      is.finite(working_df$ratio) & working_df$ratio!=0,
    ],
    start=list(
        a=.001,
        b=.012,
        c=-30
      )
    ),
    error = function(e) {
      skip <- T
    }
  )
  if(skip) {
    next
  }
  mod_sum <-summary(work_fit)
  us_agg$amp[rownum] <- coef(work_fit)['a']
  us_agg$amp_err[rownum] <- mod_sum$coefficients[,2]['a']
  us_agg$off[rownum] <- coef(work_fit)['c']%%365
  us_agg$off_err[rownum] <- mod_sum$coefficients[,2]['c']
  # na.omit(working_df)
}
```
```{r eval=F}
us_agg <- na.omit(us_agg)
us_agg$amp[us_agg$amp <0 & us_agg$off >= 365/2 ] <- -1*us_agg$amp
us_agg$off[us_agg$amp <0 & us_agg$off >= 365/2 ] <- us_agg$off-365/2
```
```{r}
us_agg <- mutate(us_agg,amp_max=amp+amp_err) %>%mutate(amp_min=amp-amp_err)
```
```{r}
us_lat_plot <- ggplot(
  us_agg,
  mapping =aes(x=latitude,y=abs(amp)))+
  coord_cartesian(xlim=c(12,50),ylim=c(0,.01))+
  geom_point(alpha=.01)
# +geom_errorbar(mapping=aes(x=Latitude,))
```

## Correlation latitude of counties to their periodic amplitude
```{r}
us_lat_plot
```

# Conclusion
It doesn't appear there is a relation between periodic changes in mortality and latitude based on my model.
This is encouraged by the lack of correlation in the offset. If there was a shared seasonal periodicity, there would be similarity across latitude of the offset of the period.
```{r}
ggplot(
     us_agg,
     mapping =aes(x=latitude,y=off))+
     geom_point(alpha=.1)
```

 ## Methodological Assumptions

- Period of 1 year
- Skip past 0 deaths

# Further questions
Consider state and country analysis.

## Bias and Mitigation
# TODO
Suppress warnings
https://stackoverflow.com/questions/13286531/how-to-suppress-warnings-when-plotting-with-ggplot